---
title: ggplot2 advanced
author: Bala Kuppusamy
date: '2019-02-23'
weight: 10
slug: ggplot2-advanced
categories:
  - R
tags:
  - plot
  - R Markdown
cover: /images/ggridges.png
output:
  blogdown::html_page:
    toc: true
    fig_width: 6
    dev: "svg"
---

# Analyzing CMPD Traffic Stops

For this problem set, we'll examine a data set of stops by the Charlotte-Mecklenburg Police Department (CMPD). 

Our focus will be to understand what factors are related to whether someone is searched or not for a traffic stop.

For our data set, you'll load the csv file we saved in the setup. This should be in your data folder.

```{r load-packages-data, warning=FALSE, message=FALSE}
library(tidyverse)
library(scales)
library(ggspatial) # make sure to install if you don't have it

data_folder <- './data'
csv_file <- paste0(data_folder, '/Officer_Traffic_Stops.csv')
shp_file <- paste0(data_folder, '/CMPD_Police_Divisions/CMPD_Police_Divisions.shp')

```

``` {r load-csv-file}

df <- read_csv(csv_file)

```

## Demographics of drivers (1 point)

First, look at the data using the `glimpse()` function from `dplyr`

```{r glimpse}
glimpse(df)
```

Notice the different variable types: character (chr), num (numeric), and datetime (POSIXct).

Let's consider our target variable: `Was_a_Search_Conducted`.

**Plot** a bar chart that counts the number of records by `Was_a_Search_Conducted`.

```{r bar}

p <- ggplot(data = df, mapping = aes(x = Was_a_Search_Conducted, fill = Was_a_Search_Conducted))
p + geom_bar() + 
  geom_text(stat = 'count', mapping = aes(label = ..count..), color = 'purple') +
  guides(fill = FALSE) +
  labs(title = 'Count of records based on whether a search was conducted', x = 'Search conducted?', y = 'Count of records')

```

**It's okay if students use `geom_col()` instead of `geom_bar()`**

How well balanced is the data set by this field? 
**The data set is NOT balanced by this field; about 95% of records are having a No value; only about 5% records have a Yes value.**

Next, let's consider the age range of the driver. 

**Plot** a histogram of `Driver_Age`. Determine an appropriate number of bins.

**Since the histogram is based on age, the appropriate `binwidth` seems to be either 5 or 10.**

```{r histogram}
p <- ggplot(data = df, mapping = aes(x = Driver_Age))
p + geom_histogram(binwidth = 5) +
  labs(title = 'Distribution of records based on driver age', x = 'Driver age', y = 'Count of records')
p + geom_histogram(binwidth = 10) +
  labs(title = 'Distribution of records based on driver age', x = 'Driver age', y = 'Count of records')
```

Once you go above (around) 40-50 bins, you'll notice some points stick out. 

What is happening? 
**Looks like the age data is not very accurate; good portion of the age information seems to have been rounded off to the nearest 5. In the below data, the number of records at the intervals of 5 (20, 25, 30, 35, 40, 45, 50) seems to be disproportionately higher compared to number of records for other ages. Here are the ages with top population count: **

```{r}
df %>%
  group_by(Driver_Age) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

```


**Plot** a density plot of `Driver_Age`. Add in a fill to be "lightblue". Determine an appropriate kernel density to use (`adjust`).

```{r density}
p <- ggplot(data = df, mapping = aes(x = Driver_Age))
p + geom_density(fill = 'lightblue', color = 'skyblue3', adjust = 0.8) +
  labs(title = 'Distribution of records based on driver age', x = 'Driver age', y = 'Density of records')
```

**Plot** a box plot with `Was_a_Search_Conducted` on the x-axis and `Driver_Age` on the y-axis.

```{r boxplot}
p <- ggplot(data = df, mapping = aes(x = Was_a_Search_Conducted, y = Driver_Age))
p + geom_boxplot() +
    labs(title = 'Distribution of driver age based on search conducted flag', x = 'Search conducted?', y = 'Driver age')
```

**Plot** a violin plot.

```{r violin}
p <- ggplot(data = df, 
            mapping = aes(x = Was_a_Search_Conducted, y = Driver_Age, fill = Was_a_Search_Conducted, color = Was_a_Search_Conducted))
p + geom_violin() +
  guides(fill = FALSE, color = FALSE) +
  labs(title = 'Distribution of driver age based on search conducted flag', x = 'Search conducted?', y = 'Driver age')
```

From the plots above, do you think the age of the driver is a significant factor in whether a search was conducted? Why or why not?
**Yes, I think age of the driver is a significant factor in whether search was conducted. The median age of drivers where search conducted is at least 5 years less compared to the search-not-conducted population; the entire 3rd quadrant of the search-conducted population is lesser than 35 years, which is just the median of the search-not-conducted population.**

## Date of stop (1 point)

Let's plot the number of stops by time. 

Recalling part one, the `Month_of_Stop` variable is a character, not a date variable. The datatime's are simply when the data was collected; not when the stop occurred. Therefore, we'll need to convert the `Month_of_Stop` variable from a character to a Date format.

```{r}
glimpse(df)
```


Let's first cleanup the date field using `tidyverse` packages like `stringr` and `lubridate`.

```{r date-cleanup}
library(stringr); library(lubridate)

df <- df %>%
  mutate(Month_of_Stop = str_replace_all(Month_of_Stop, "/","-")) %>% # replace "/" with "-"
  mutate(Month_of_Stop = paste0(df$Month_of_Stop,"-01")) %>% # add in day
  mutate(Date = ymd(Month_of_Stop)) # created a date field
```

**Plot** a line chart with the number of traffic stops for each month (hint: start with the `count()` function by Date then feed into ggplot. Remember the count variable is named 'n'.).

```{r stops}
by_date <- df %>%
  count(Date)

p <- ggplot(data = by_date, mapping = aes(x = Date, y = n))
p + geom_line(color = 'skyblue3', size = 1.2 ) +
  geom_point() +
  scale_x_date(labels = date_format('%b %Y'), date_breaks = '4 month') +
  labs(title = 'Distribution of traffic stops by month', x = 'Month', y = '# of traffic stops')
  # theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

What is the trend (i.e., long term rate of change) of the number of traffic stops in Charlotte?
**The traffic stops have increased year-over-year between 2016 vs 2017 for the same periods. Also, the traffic stops seem to have some seasonality, whereby they peak between months of Jan-April, & then gradually reduce towards end of the year.**

**Plot** the same plot but add in `facet_wrap()` by the `Reason_for_Stop` variable.

```{r stops-by-reason, fig.width=8, fig.height=5}
by_date <- df %>%
  count(Date, Reason_for_Stop)

p <- ggplot(data = by_date, mapping = aes(x = Date, y = n))
p <- p + geom_line(color = 'skyblue3', size = 1.2 ) +
  scale_x_date(labels = date_format('%b %Y'), date_breaks = '4 month') +
  labs(title = 'Distribution of traffic stops by month, split by reason', x = 'Month', y = '# of traffic stops') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
p + facet_wrap(~Reason_for_Stop)
```

What is a problem with this plot?
**The plot indicates that majority of the variations are caused by 'Vehicle Regulatory' & 'Speeding' categories. The actual values & variations for other reasons is not clearly visible since they have very low values & the scales for all the charts are set as fixed.**

To address this problem, you will need to figure out how to adjust the scale. To do this, you need to use R's documentation to see whether there is a parameter in `facet_wrap`.

Go to your RStudio console and type `?facet_wrap`.

What parameter allows you to modify the scales of `facet_wrap`? `scales = 'free_y'`

**Plot** the same plot but with a free y-axis scale.

```{r stops-by-reason-2, fig.width=8, fig.height=5}
p + facet_wrap(~Reason_for_Stop, scales = 'free_y')

```

Which type of police stop has had the most volatility (i.e., big swings in number of stops)?
**Vehicle Regulatory, Speeding.**

What is one problem with allowing the y-axis be free?
**Even minor variations are displayed similar to the ones having bigger swings; challenging to differentiate high impact variations vs low impact variations.**

Small multiples tends to be less effective when each of the variables are on different scales or magnitudes.

Let's consider instead CMPD traffic stops but by CMPD division. These are more even spread by division than the type of stop.

**Plot** a line chart (optional points too) for stops by `Date` (x axis) and counts ('n', or whatever you named your count variable) (y axis). (hint: to modify how the date is shown, use the layer `scale_x_date(date_labels = "%Y") +` to show only the year. Feel free to modify by looking at `?scale_x_date`.)

```{r stops-by-division}
by_division <- df %>%
  count(Date, CMPD_Division)

p <- ggplot(data = by_division, mapping = aes(x = Date, y = n))
p <- p + geom_line(mapping = aes(color = CMPD_Division), size = 1.2) +
  geom_point(alpha = 0.8, size = 0.8) +
  facet_wrap(~CMPD_Division) +
  guides(color = FALSE) +
  scale_x_date(labels = date_format('%b %Y'), date_breaks = '4 month') +
  labs(title = 'Distribution of traffic stops by month, split by division', x = 'Month', y = '# of traffic stops') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
p
```

What are three observations you can make about the number of police stops by divison? (hint: just write about what's in the data.)

1. **The number of traffic stops are evenly spread-out across all major divisions - range of 200-500 stops per division.**

2. **The number of traffic stops show a considerable increase in 2017 vs 2016 in these divisions - Eastway, North Tryon.**

3. **Divisions - South, Westover, show a decreasing trend in 2017 compared to 2016, indicating a possible reduction in violations.**

Next, this doesn't help tell us where these areas are. For that, let's use a shape file to create a chloropleth of stops by division.

## Geography (1 point)

For this example, we'll create a cholorpleth for the number of police stops by police division.

To do this, we need to use the `sf` package. (For help along the way, see [this tutorial](http://strimas.com/r/tidy-sf/) on `sf` package.)

```{r load-sf}
library(sf)
library(viridis)

cmpd <- st_read(shp_file)
```

Note that while we have five files, we only load in the shapefile (`.shp`) file. This is typical but know that to use this file you would need the other four files in the same folder as your shapefile.

**Plot** cmpd using the `geom_sf` package where you provide `fill = DNAME` as the only aesthetic. Add in a title saying "CMPD Divisions" and add the `theme_bw()` theme to make translate the file into the black and white template.

```{r geom_sf}
p <- ggplot(data = cmpd, mapping = aes(fill = DNAME))
p + geom_sf() +
  scale_fill_discrete('Division Name') +
  theme_bw() +
  labs(title = 'CMPD divisions')
```

One problem with this map is it's hard to read the division names. That is, it may be better to remove the legend and put the labels of each division within the plot.

To do this, we can use the related `geom_sf_label()` geom, using the name of the division as the aesthetic label.

**Plot** the same plot from above but with the name of the division as the label. 

1. You'll likely need to reduce the size of the label, using the `size` paramater. You should likely set the `size` to under 2. 

2. Make sure to remove the legend (it's redundant and no longer necessary).

3. Create a new variable named `Name` that removes the term " Division". This term is redundant and takes up a lot of space in the labels from `DNAME`. To do this step, use this snippet of code at the top of your pipeline:

```{r warning=FALSE}
cmpd <- cmpd %>%
  mutate(Name = as.character(DNAME)) %>%
  mutate(Name = str_replace_all(Name, " Division",""))

p <- ggplot(data = cmpd, mapping = aes(fill = Name))
g <- p + geom_sf() +
  geom_sf_label(mapping = aes(label = Name), color = 'white', size = 2) +
  guides(fill = FALSE) + 
  theme_bw()
```

4. Save it as an object named `g`. Make sure to call it once so that the map will output.

```{r geom-labels, warning=FALSE}
g
```

**Plot**: Now let's change the projection type. Simply add in `coord_sf(crs = st_crs(102003))` as an additional layer to `g`.

```{r albers}
g + coord_sf(crs = st_crs(102003))
```


Do you notice any difference between the two maps?
**Yes, the lattitude & longitude are displayed in an angled direction (indicating Albers projection).**

## Advanced Plot: (0.5 point)

Now, let's create a chloropleth. Below is the code to create an advanced plot.

In this problem, you need to explain what each step below is doing:

- `mutate()`: **Adds a new attribute 'CMPD_Division' to 'cmpd' dataframe, by converting 'DNAME' attribute to characters.**
- `inner_join()`: 
    1. **Summarizes the 'df' dataframe by 'CMPD_Division' & 'Date' to get the count of traffic stops.**
    2. **Joins the 2 dataframes by matching 'CMPD_Division'; add 2 additional attributes 'Date' & 'n' to 'cmpd' dataframe.**
- `mutate()`: **Extracts 'year' from 'Date' attribute using libridate library function; adds the 'year' attribute to 'cmpd' dataframe.**
- `geom_sf()`: **Plots the shape from cmpd shape file, with the fill aesthetic based on the number of traffic stops.**
- `scale_fill_viridis()`: **Uses viridis color scale for fill aesthetics; specifying title for the legend & values to have thousand seperator.**
- `labs()`: **Specifies different labels for the plot - chart title (shows at the top) and caption (shows at the bottom right).**
- `annotation_scale()`: **Produces a small annotation (reference scale) indicating spatial distances.**
- `facet_wrap()`: **Splits the plot by year producing 2 seperate plots for 2016 vs 2017.**
- `theme_bw()`: **Uses black & white theme; have a white background for the plot instead of the default grey.**
- `theme()`: (what are each of the options doing in `theme()`?)
    1. `legend.position` - **Indicates the legend display position - bottom of the plot, in this case.**
    2. `plot.title` - **Controls the title of the plot - specifies title formatting inforamtion like the font face & size.**
    3. `axis.text` - **Controls how the text in x & y axis are displayed; blanks them out in this case.**
    4. `axis.ticks` - **Controls how the tick marks in x & y axis are displayed; blanks them out in this case.**
- `ggsave()`: **Saves the plot into physical files in the file system; saves it as a pdf & a png in this case.**

```{r adv-plot}
cmpd_chloropleth <- cmpd %>% 
  mutate(CMPD_Division = as.character(DNAME)) %>% 
  inner_join(count(df, CMPD_Division, Date), by = "CMPD_Division") %>%
  mutate(Year = lubridate::year(Date)) %>%
  ggplot() +
  geom_sf(aes(fill = n)) +
  scale_fill_viridis("Traffic Stops", labels = scales::comma) +
  labs(title = "CMPD Traffic stops by CMPD Division",
       caption = "Source: CMPD") +
  annotation_scale(location = "bl", width_hint = 0.2) +
  facet_wrap(~Year) +  
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = rel(1.5)),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank())

cmpd_chloropleth

# ggsave(cmpd_chloropleth, filename = "cmpd_chloropleth.pdf",
#       width = 7, height = 5, units = "in")
# ggsave(cmpd_chloropleth, filename = "cmpd_chloropleth.png",
#       width = 7, height = 5, units = "in")
```


## Choosing a ggplot extensions  (1 point)

Go to [ggextensions website](https://www.ggplot2-exts.org/). Then click Galleries to explore the different ggplot extensions. Scroll through and see if any catch your eye.

Now, select **one** of the ggextension libraries below and install the package (through CRAN):

- [`ggridges`](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) / [example to recreate](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html#varying-fill-colors-along-the-x-axis)^[Run both plots. Make sure to install the `viridis` package or else you'll get an error!]

- [`ggalt`](https://github.com/hrbrmstr/ggalt) / [example to recreate](https://github.com/hrbrmstr/ggalt#horzon-chart)^[Make sure to install `hrbrthemes`!]

- [`ggrepel`](https://github.com/slowkow/ggrepel) / [example to recreate](https://github.com/slowkow/ggrepel/blob/master/vignettes/ggrepel.md#align-labels-on-the-left-or-right-edge)

- [`ggstatsplot`](https://github.com/IndrajeetPatil/ggstatsplot) / [example to recreate](https://github.com/IndrajeetPatil/ggstatsplot#ggscatterstats)^[Run all three examples in the `ggscatterstats` section.]

**Plot** the related example

```{r ggextension-examples, warning=FALSE}
# library(ggalt)
# library(hrbrthemes)
# library(tidyverse)
# library(ggplot2)
# library(gridExtra)
# # hrbrthemes::import_roboto_condensed()
# 
# packageVersion("ggalt")
# 
# sports <- read_tsv(file = './data/activity.tsv.txt')
# 
# sports <- sports %>%
#   group_by(activity) %>%
#   filter(max(p) > 3e-04, !grepl('n\\.e\\.c', activity)) %>%
#   arrange(time) %>%
#   mutate(p_peak = p / max(p), 
#          p_smooth = (lag(p_peak) + p_peak + lead(p_peak)) / 3,
#          p_smooth = coalesce(p_smooth, p_peak)) %>%
#   ungroup() %>%
#   do({
#     rbind(., filter(., time == 0) %>% mutate(time = 24*60))
#   }) %>%
#   mutate(time = ifelse(time < 3 * 60, time + 24 * 60, time)) %>%
#   mutate(activity = reorder(activity, p_peak, FUN=which.max)) %>%
#   arrange(activity) %>%
#   mutate(activity.f = reorder(as.character(activity), desc(activity))) %>%
#   mutate(time2 = time / 60)
# 
# p <- ggplot(data = sports, mapping = aes(x = time2, y = p_smooth))
# p + geom_horizon(bandwidth=0.1)

library(viridis)
library(ggridges)

d <- data.frame(x = rep(1:5, 3) + c(rep(0, 5), rep(0.3, 5), rep(0.6, 5)),
                y = c(rep(0, 5), rep(1, 5), rep(3, 5)),
                height = c(0, 1, 3, 4, 0, 1, 2, 3, 5, 4, 0, 5, 4, 4, 1))

ggplot(d, aes(x, y, height = height, group = y, fill = factor(x + y))) +
  geom_ridgeline_gradient() +
  scale_fill_viridis(discrete = TRUE, direction = -1, guide = "none")

```

Now, with the same package you ran, make a **plot** with that package and the gapminder data. You can choose any of the data frames (i.e., years). Make sure your plot has at least six functions (e.g., ggplot() + geom_point() is two functions and `dplyr` functions count as well.)

```{r ggextension-custom, warning=FALSE, message=FALSE}
library(gapminder)

p <- gapminder %>%
  mutate(decade = year - (year %% 10) + 10) %>%
  group_by(continent, country, decade) %>%
  summarise(lifeExp = mean(lifeExp)) %>%
  ggplot(mapping = aes(x = lifeExp, y = continent, color = continent, fill = continent)) +
  geom_density_ridges(rel_min_height = 0.01, alpha = 0.4) +
  guides(fill = FALSE, color = FALSE) +
  facet_wrap(~decade) +
  labs(title = 'Life expectancy distribution in continents, spread across decades', x = 'Life Expectancy', y = 'Continent')
p
```

Describe what you have found using that plot (write at least 3 sentences): 
**The contries in Europe had been concentrated towards the higher end of the Life Expectancy sprectrum since 1960s. Whereas, countries in other continents - Asia, Americas, & Africa, were concentrated towards the lower end in 1960s & have slowly moved over to the higher end of the spectrum by 2010. The rate of movement is high for countries in Americas & Asia, & it is very slow for Africa.**

## Intro to interactivity (0.5 points)

For even more fun, **plot** an interactive HTML plot using the code for any of the plots above (fair warning, some of the ggextensions may not work well). 

The easiest way to do this is to use the `plotly` package (install it with the "Packages" panel in RStudio), and then to use its `ggplotly()` function.

I've given you some commented-out code below (commented out so that R doesn't yell at you about the code not working when you knit). 

Also, check out [the documentation](https://plot.ly/ggplot2/), especially [this page](https://plot.ly/ggplot2/interactive-tooltip/) about customizing the tooltips that show up when you hover over points or areas.

```{r interactive, message=FALSE}
library(plotly)
library(gapminder)

p2 <- ggplot(data = gapminder, mapping = aes(x = lifeExp, group = continent, color = continent, fill = continent)) +
  geom_density(alpha = 0.4)
  # facet_wrap(~ continent)

 
# ggplotly(p = p2, tooltip = 'x', displayModBar = TRUE, modeBarButtonsToRemove = c('pan2d','lasso2d'))
ggplotly(p = p2, tooltip = 'x')
```
